```{r setupdat055, echo=FALSE}
# Read in the dat1 file created from 03-randomizeddesigns.Rmd
if(!exists("dat1")){ dat1 <- read.csv(here::here('dat1_with_designs.csv')) }

dat1$rankY <- rank(dat1$Y)
ndat1 <- nrow(dat1)

dat1$y1s <- dat1$y1 / sd(dat1$y1)
dat1$y0s <- dat1$y0 / sd(dat1$y0)

trueates <- mean(dat1$y1s) - mean(dat1$y0s)
trueate <- mean(dat1$y1) - mean(dat1$y0)
```

# Power Analysis {#poweranalysis}

In this chapter, we provide examples of how we assess statistical power for different experimental research designs. We often prefer to use simulation to assess the power of different research designs because we rarely have designs that fit easily into the assumptions made by analytic tools.

By "analytic tools," we refer to methods of calculating power (or a minimum required sample size, etc.) based on mathematical derivations under particular assumptions. For instance, in an *i.i.d.* two-arm design with random assignment of half the sample, no covariates, a roughly normally distributed outcome, and equal variance in each treatment group, it's possible show that we would have 80% power to estimate a difference in means of $\Delta$ if we collect data on approximately $n = (5.6 \sigma /\Delta)^{2}$ observations, where $\sigma$ is the overall standard deviation of the outcome.^[See @gelman2006data, section 20.3, page 443.] But we frequently consider situations where such derivations are not readily available.

<!-- I'm not sure this introductory example is useful, in that it doesn't show what we claimed to want (that analytical and simulation yield the same results in a simple example; I cut this text [Bill] in my recent edit) due to the skewed nature of this outcome. For the purposes of this walkthrough, I think skew is a complication we want to put aside? And also, both designs have power of basically 100% with this example data, so the description in the text is no longer accurate. So let's just jump to the more detailed examples of each approach below?  -->

<!-- Imagine that we thought a study would have an effect of about 1 standard deviation -- this is more or less the effect difference in the example data we've been using. How much statistical power would we have with a sample size of only 40 observations? -->

```{r, echo = F, eval = F}
## Skipping for now
## Set up some DeclareDesign objects for our simulation
population <- declare_population(dat1)
potentials <- declare_potential_outcomes(Y ~ y0 + Z * delta)
assignment <- declare_assignment(Z=complete_ra(N))
reveal <- declare_reveal(Y, Z)
design <- population + potentials + assignment + reveal

## Draw one simulated dataset
simdat1 <- draw_data(design)

## Set inquiry and estiamtor
estimand <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))
estimator <- declare_estimator(Y ~ Z, inquiry = estimand, label = "Simple D-I-M")

## Illustrate how the estimator works
estimator(simdat1)

## Add the inquiry and estimator to our design
designPlusEst <- design + estimand + estimator
```

<!-- Now that the setup is complete, we can assess the statistical power of our -->
<!-- proposed design with $N=100$ and an effect of roughly 1 SD. The output below -->
<!-- shows the statistical power as well as the false positive rate (called -->
<!-- "Coverage" below), as well as the difference between the mean difference we -->
<!-- calculate and the true average treatment effect (called "Bias" below). -->

```{r diagpow1, cache=TRUE, echo = F, eval = F}
## Skipping for now
sim_power <- diagnose_design(designPlusEst, sims = 200)
```

<!-- Initially, this analytic approach suggests more power than the simulation based approach -- a difference that arises from our fairly skewed outcome. -->

```{r powerttest, echo = F, eval = F}
## skipping for now
analytic_power <- power.t.test(n = 100, delta = delta, sd = sd(dat1$y0))
analytic_power
```

<!-- For ways to compare different sample sizes, effect sizes, etc. see [more -->
<!-- information from the DeclareDesign -->
<!-- package](https://declaredesign.org/blog/2018-10-02-power-strategies.html). -->

## An example of the off-the-shelf approach

To demonstrate how analytical power analysis works in principle, consider the R function `power.t.test()`. This can be used for power calculations in designs where a two-sample t-test is an appropriate estimation strategy (with no adjustment for blocking, clustering, or covariates).

When using this function, there are three parameters that we're most concerned
with, two of which must be specified by the user. The third is then calculated and returned by the function. These are:

* `n` = sample size, or number of observations *per treatment group*
* `delta` = the target effect size, or a minimum detectable effect (MDE)
* `power` = the probability of detecting an effect if in fact there is a true effect of size `delta`

Note that there is also the parameter `sd`, representing the standard deviation of the outcome. This is set to 1 by default unless `power.t.test()` is instructed otherwise.

Say, for example, you want to know the MDE for a two-arm study with 1,000
participants, of which half are assigned to treatment. Using `power.t.test()` you would specify:

```{r powerttest2}
power.t.test(
  n = 500, # The number of observations per treatment arm
  power = 0.8 # The traditional power threshold of 80%
  )
```

If all we wanted to extract was the MDE, we could instead write:

```{r}
power.t.test(n = 500, power = 0.8)$delta
```

If we request the sample size instead, we can illustrate that this is applying an expression like the one we mention above: $$n = (5.6/0.1773605)^{2} \approx 1000$$

And now via R code:

```{r}
power.t.test(delta = 0.1773605, power = 0.8)$n * 2
```

If you need to, you can adjust other parameters, like the standard deviation of the outcome, the level of the test, or whether the test is one-sided rather than two-sided. There are also other functions available for different types of outcomes. For example, if you have a binary response, you can use `power.prop.test()` to calculate power for a similar kind of simple difference in proportions test.

An equivalent approach in Stata is as follows:

```
power twomeans 0, power(0.8) n(1000) sd(1)
```

Stata users can learn more about available tools by checking out Stata’s
plethora of [relevant help files](https://www.stata.com/features/power-and-sample-size/).

## An example of the simulation approach

We can compare the output of `power.t.test()` to the output from a
simulation-based (i.e., computational) approach, which we illustrate in the code chunk below. Notice that our example below relies on manually-written functions that could be copied and adapted by OES team members to the needs of different projects. This code is more detailed than we need for simple cases, but it provides useful flexibility for simulating more complicated designs (particularly designs we cannot specify as easily within the `DeclareDesign` framework).

```{r poweranalysistools}
## OES Power Simulation Toolkit:
## Replicate, Estimate, Evaluate (REE)
## 
## replicate_design(...) ---: Generate replicates of a simulated dataset
## estimate(...) -----------: Estimate the null test-stat for treatment(s).
## evaluate_power(...) -----: Evaluate power to detect non-zero effects.
## evaluate_mde(...) -------: Find MDE, searching over range of effect sizes.
## evaluate_bias(...) ------: Compute bias and other diagnostics.

## Required packages
require(magrittr)
require(fabricatr)
require(foreach)

##### REPLICATE a hypothetical design R times #####

# Inputs:
# - (1) number of replicates desired
# - (2) additional arguments that are passed to fabricate() (see the ...).
#   - Generally, each arg is a separate variable to generate.
#   - Later vars can be a function of earlier ones.
#   - See our examples below, it's simpler than it sounds!
replicate_design <- 
  
  function(R = 200, ...) {
    
    # Function: produce one draw of the simulated dataset/design
    design <- function() {
      fabricatr::fabricate(
        ...
        ) %>%
        list
      }    
    
    # Use replicate() to replicate that design R times
    rep <- replicate(
      n = R,
      expr = design()
      )
    
    # Output will be a list of dataframes.
    # For each, add a variable indicating which sim# it is
    for(i in 1:length(rep)) {
      rep[[i]] <- rep[[i]] %>%
        dplyr::mutate(
          sim = i
        )
      }
    
    return(rep)
    
  }

##### ESTIMATE results using those replicated dfs #####

# Inputs:
# - (1) Estimation formula (y ~ x1 + x2 + ...)
# - (2) Variables(s) we want to be powered to estimate the effects of
#     - Generally just the treatment var(s)
# - (3) Data the estimator should be applied to (list of dfs)
# - (4) The estimator (the default is OLS with HC2 errors)
estimate <- function(
  formula,
  vars,
  data = NULL,
  estimator = estimatr::lm_robust
  ) {

  # Pass the list of dfs to map().
  # map() applies the procedure specified below to each df in the list.
  data %>%
    
    purrr::map(
      
      # For each dataframe in the list, apply the specified estimator,
      # using the specified formula.
      ~ estimator(
          formula,
          data = .
          ) %>%
        
        # tidy up the results and specify the rows of estimates to keep
        estimatr::tidy() %>%
          dplyr::filter(
            .data$term %in% vars
            ) 
      
      ) %>%
    
    # Append the results from each sim replicate into a single dataframe
    dplyr::bind_rows() %>%
    
    # Add some more useful labels
    dplyr::mutate(
      sim = rep(1:length(data), each = n() / length(data)),
      term = factor(.data$term, levels = vars)
      )
  
}

##### EVALUATE power of the design #####

# Inputs:
# - (1) Results produced by estimate() above
# - (2) Hypothetical effects we want power estimates for
# - (3) Desired alpha (significance) level
evaluate_power <- function(data, delta, level = 0.05) {
  
  # Make sure delta (may be scalar or vector) was specified
  if (missing(delta)) {
    stop("Specify 'delta' to proceed.")
  }
  
  # Apply the following (i.e., after %do%) to each delta separately,
  # appending the results with bind_rows at the end.
  foreach::foreach(
    i = 1:length(delta),
    .combine = "bind_rows"
    ) %do% {
      
      # Start with the df of estimates
      data %>%
        
        # Create variables storing the relevant delta and new test stat
        dplyr::mutate(
          delta = delta[i],
          new_statistic = (.data$estimate + .data$delta) / .data$std.error
          ) %>%
        
        # Similar to group_by, result here is list of dfs for each term
        dplyr::group_split(.data$term) %>%
        
        # Separately for the df for each term, get p for each replicate
        purrr::map(
          ~ {
            tibble::tibble(
              term = .$term,
              delta = .$delta,
              p.value = foreach(
                j = 1:length(.$new_statistic),
                .combine = "c"
                ) %do% mean(abs(.$statistic) >= abs(.$new_statistic[j]))
              )
            }
          )
      } %>%
    
    # Organize by term and delta
    group_by(.data$term, .data$delta) %>%
    
    # Average over repliacates to get power for each term/delta combination
    summarize(
      power = mean(.data$p.value <= level),
      .groups = "drop"
      )
  
}

##### EVALUATE the min. detectable effect #####
# Helps summarize the results of evaluate_power() above,
# basically a wrapper for evaluate_power()

# Inputs:
# - (1) Results produced by estimate() above
# - (2) Range of hypothetical effects we want to consider (delta above)
# - (3) How fine-grained do we want changes in delta to be?
# - (4) Alpha (significance) level
# - (5) Minimum power we want to accept
evaluate_mde <- function(
  data, 
  delta_range = c(0, 1),
  how_granular = 0.01,
  level = 0.05,
  min_power = 0.8
  ) {
  
  # Use the function designed above to get power estimates
  eval <- evaluate_power(
    data = data, 
    delta = seq(delta_range[1], delta_range[2], how_granular),
    level = level
    ) %>%
    
    # Organize data by term
    dplyr::group_by(
      .data$term
      ) %>%
    
    # Get the MDE at our desired power level for each term
    dplyr::summarize(
      MDE = min(.data$delta[.data$power >= min_power]),
      .groups = "drop"
      )
  
  return(eval)
  
}

##### EVALUATE Bias #####
# Helps summarize the results of estimate() above.
# Pass results of estimate() to this function.

# Inputs:
# - (1) Data produced by estimate() above
# - (2) True ATE
evaluate_bias <- function(
  data, 
  ATE = 0
  ) {
  
  # Start with the estimates for each replicated dataset
  smry <- data %>%
    
    # Add a variable representing the true ATE
    dplyr::mutate(
      ATE = rep(ATE, len = n())
      ) %>%
    
    # Organize estimates by term
    dplyr::group_by(
      .data$term
      ) %>%
    
    # Summarize across replicates, within each term
    dplyr::summarize(
      "True ATE" = unique(.data$ATE),
      "Mean Estimate" = mean(.data$estimate),
      Bias = mean(.data$estimate - .data$ATE),
      MSE = mean((.data$estimate - .data$ATE)^2),
      Coverage = mean(
        .data$conf.low <= .data$ATE & .data$conf.high >= .data$ATE
        ),
      "SD of Estimates" = sd(.data$estimate),
      "Mean SE" = mean(.data$std.error),
      Power = mean(.data$p.value <= 0.05),
      .groups = "drop"
    )
  
  return(smry)
  
}
```

Results are shown in the subsequent figure. Though the computational estimates are slightly different, they comport quite well with the analytic estimates.

```{r powersimvsanalytic}
## Parameters used for both sets of calculations
n <- 1000 # Sample size
d <- 0.2 # Effect size to consider

## Analytical power estimates
power_data <-
  
  tibble(
    d = seq(0, 0.5, len = 200),
    power = power.t.test(n = n / 2, delta = d)$power
    )

## Save initial plot; add simulation results below
g <- ggplot(power_data) +
  
  geom_line(aes(d, power, linetype = "power.t.test()")) +
  
  labs(
    x = expression(delta),
    y = "Power",
    title = "Power for Simple Difference in Means Test"
    ) +
  
  scale_y_continuous(
    n.breaks = 6
    ) +
  
  geom_hline(
    yintercept = 0.8,
    col = "grey25",
    alpha = 08
    ) +
  
  ggridges::theme_ridges(
    center_axis_labels = TRUE,
    font_size = 10
    ) 

## Comutational power estimates, using the functions above
sim_power_data <-
  
  replicate_design(
    N = n,
    y = rnorm(N),
    x = randomizr::complete_ra(
      N, m = N / 2
      )
    ) %>%
  
  estimate(
    form = y ~ x, vars = "x"
    ) %>%
  
  evaluate_power(
    delta = seq(0, 0.5, len = 200)
    )

## Add results from the simulation to the plot and compare
g + 
  
  geom_line(
    data = sim_power_data,
    aes(delta, power, linetype = "simulation"),
    color = "grey25"
    ) +
  
  labs(
    linetype = "Method:"
    ) +
  
  theme(
    legend.position = "bottom"
    ) 
```

As mentioned above, we produced those computational estimates using some pre-written functions that are currently housed in OES's GitHub code library. These functions are also laid out and explained step-by-step in comments in a code chunk above.

These tools are designed around a simple workflow, and they should help remove some of the programming that may otherwise be a barrier to project teams calculating power computationally. The workflow proceeds as follows:

1. Replicate
2. Estimate  
3. Evaluate
  
The first step, **Replicate**, entails specifying an example data-generation process (which may include only an outcome variable and treatment assignment) and simulating it multiple times to create a series of randomly generated datasets. Each individual dataset produced is a *sample replicate*.

The next step, **Estimate**, entails estimating effects for select treatments within each sample replicate. We can use those estimates to produce a distribution of test statistics for each effect size of interest.

Finally, the last step, **Evaluate**, entails using those test statistics to evaluate our power to detect a range of different effect sizes.

This workflow is supported by three functions: `replicate_design()`,
`estimate()`, and `evaluate_power()`. Here's the simulation code used to generate Figure 1, in more detail:

```{r, eval=FALSE}
## 1. Replicate:

# Output is a list of dfs
rep <- replicate_design(
  R = 200,   # Number of sample replicates
  N = 1000,  # Sample size of each replicate
  y = rnorm(N), # Normally distributed response
  x = rbinom(N, 1, 0.5) # Binary treatment indicator
  ) 

## 2. Estimate:

# Output is a dataframe of estimates from each sample replicate
est <- estimate(
  y ~ x, # Regression formula
  vars = "x", # Treatment variable(s)
  data = rep # Sample replicates
  ) 

## 3. Evaluate:

# Output is a list of dfs
pwr_eval_sim <- evaluate_power(
  data = est, # Estimates, from estimate() above
  delta = seq(0, 0.5, len = 200) # Effect sizes to consider
  )   
```

The final product, `pwr_eval_sim` above, reports the power for each of the user-specified effect sizes (`delta`) and (`vars`) model terms specified when calling `estimate()`. The output can be used to plot power curves or to compute minimum detectable effects.

These functions help make the process of performing computational power analysis for OES projects easier, while still providing ample room for flexibility in both design and estimation strategy. For example, `replicate_design()` is a wrapper for `fabricate()` in the [fabricatr](https://declaredesign.org/r/fabricatr/) package. This gives users the ability to generate multi-level or nested data-generating processes, specify additional covariates, or determine whether treatment randomization is done within blocks or by clusters. By default, estimates are returned using `lm_robust()` from the [estimatr](https://declaredesign.org/r/estimatr/) package, but alternative estimators can be specified. Say, for example, you have a binary response and a set of covariates, and your design calls for using logistic regression. You could generate estimates for such a design as follows:

```{r, eval=FALSE}
## Define logit estimator function
logit <- function(...){ glm(..., family = binomial)}

## Pass this to the estimate() function above
est <- estimate(
  y ~ x + z1 + z2, data = rep, estimator = logit
  ) 
```

Other tools for power simulation exist as well. For instance, throughout this SOP, we have used [DeclareDesign](https://declaredesign.org/r/declaredesign/articles/DeclareDesign_101.html) to simulate hypothetical research designs and compare their performance. And there is no shortage of further simulation
examples that can be found online for more specialized use-cases.

## When to use which approach

For a simple difference in means test, the programming required for an analytical power analysis is much much less involved. When is computational power analysis worth the extra time investment?

To start, in cases where we’re interested in the power to detect a simple difference in means, or a difference in proportions for binary responses, it is probably sufficient to use `power.t.test()` (for means) or `power.prop.test()` (for proportions).

However, OES projects often involve design features or analytic strategies that are difficult to account for using off-the-shelf tools. For example, we often include covariates in our statistical models to enhance the precision of our treatment effect estimates. If the gain in precision is small, then it might not be important to account for this in power calculations in the design phase of the project. But if we expect a substantial gain in precision due to including covariates, then we probably want to account for estimating power. The natural way to do this is by simulation, including the covariates in the "replicate" and "estimate" steps above. Accounting for covariates is especially useful if we can use real historical or pre-treatment data that represent the correlations we expect to see between covariates and outcomes in our later analysis of the trial data.

More complex design features or analytic strategies may make investing in the
simulation approach even more worthwhile, or downright necessary. Examples
include heterogeneity in treatment effects, a multi-arm or factorial design, or block randomization with differing probabilities of treatment between blocks -- none of which is usually easily accounted for with off-the-shelf tools. In the next section, we provide some additional examples of simulations for more complex designs or analytic strategies.

## Additional examples of the simulation approach

Here we provide two examples of research designs where simulation is well worth the extra effort. Attendant R code is included to illustrate how we could use the functions above in these cases.

### A two-by-two design with interaction

One instance where computational power analysis may be worth the investment is
in assessing power for a two-by-two factorial design with an interaction. In
such a design, the goal is to assess not only the power to detect main effects
(the average effect of each individual treatment), but also power to detect a
non-zero interaction effect between the treatments.

Say we have a design with 1,000 observations and we would like to know the
effect of two treatments on a binary outcome with a baseline of 0.25. Each
treatment is assigned to $M = 500$ individuals at random, resulting in four
roughly equal sized groups of observations after randomization: (1) a control group, (2) those assigned to treatment 1 but not treatment 2, (3) those assigned to treatment 2 but not treatment 1, and (4) those assigned to both treatment 1 and 2.

We can easily calculate power to detect the main effect of each treatment as follows:

```{r, eval=T}
two_by_two <-

  ## Basic elements of each simulated sample replicate
  replicate_design(
    N = 1000,
    y = rbinom(N, 1, 0.25),
    x1 = complete_ra(N, m = N / 2), 
    x2 = complete_ra(N, m = N / 2)
    ) %>%
  
  ## Estimate main and interaction effects
  estimate(
    form = y ~ x1 + x2 + x1:x2,
    vars = c("x1", "x2", "x1:x2")
    ) %>%
  
  ## Evaluate power
  evaluate_power(
    delta = seq(0, 0.25, len = 200)
    )
```

Using the output reported in the object `two_by_two`, we can plot the power curves for each of the main effects and the interaction effect, as shown
in Figure 2.

```{r}
ggplot(two_by_two) +
  
  ## Add a line representing power for each effect/term
  geom_line(
    aes(delta, power, linetype = term)
    ) +
  
  ## Choose linetypes that are easy to distinguish
  scale_linetype_manual(values = c("solid", "longdash", "dotted")) +
  
  ## Horizontal line for 80% power
  geom_hline(
    yintercept = 0.8,
    color = "grey25",
    alpha = 0.8
    ) +
  
  ## y-axis scale
  scale_y_continuous(
    n.breaks = 6
    ) +
  
  ## Adding labels
  labs(
    x = expression(delta),
    y = "Power",
    title = "Power for a 2x2 Design",
    linetype = "Effect for..."
    ) +
  
  ## Update some visual settings with the ridges theme
  ggridges::theme_ridges(
    font_size = 10,
    center_axis_labels = TRUE
    ) +
  
  ## Other settings (here: just legend location)
  theme(
    legend.position = "bottom"
    ) 
```

Of course, in this simple example, we could still have relied on some reasonable analytical assumptions to arrive at these estimates (see a helpful discussion [here](https://statmodeling.stat.columbia.edu/2018/03/15/need-16-times-sample-size-estimate-interaction-estimate-main-effect/)). But running a simulation saves us the trouble.

### Covariate adjustment with the Lin estimator

Another scenario where computational power analysis is worth the investment is if a design calls for covariate adjustment. This is common in OES projects, and, in many instances, the @lin_agnostic_2013 saturated regression estimator is the solution we choose.

Devising an off-the-shelf method to calculate power for such a study is
possible, but would likely require investing time doing background research to
ensure its accuracy. Alternatively, we could simply replicate, estimate, and
evaluate such a design computationally. The results will be roughly just as accurate, without regarding a review of the methods literature.

Suppose we have a sample of 1,000 observations and a continuous outcome
variable. We wish to assess the effect of some policy intervention on this
continuous outcome. Our design calls for randomly assigning only $M = 100$
individuals to receive the intervention --- perhaps because it is expensive to implement --- and the rest to control.

In addition to having data on the outcome and on treatment assignment, let's say that we also anticipate obtaining a dataset of covariates for our 1,000 observations. This data contains two variables that are prognostic of the outcome *and the treatment effect*. We’ll call these `z1` and `z2`. The first is a continuous measure and the latter is a binary indicator. Our design calls for adjusting for these covariates to improve the precision of our estimated treatment effect. We can simulate such a design to illustrate the possible benefits of covariate adjustment in terms of improved statistical power.

We begin by replicating the data-generating process:

```{r}
rep_data <-
  
  replicate_design(
    N = 1000,
    z1 = rnorm(N, sd = 3),   # Continuous covariate
    z2 = rbinom(N, 1, 0.25), # Binary covariate
    cz1 = z1 - mean(z1),     # Mean centered versions of the covariates
    cz2 = z2 - mean(z2),
    x = complete_ra(N, m = N / 10), # Randomly assign 10% to treatment
    y = ((z1 + z2) * x) + (0.8 * z1) - (1 * z2) + rnorm(N) # Simulate Y
    )
```

We then estimate and evaluate. For comparison, power is computed (1) with
covariate adjustment via the Lin estimator, (2) without covariate adjustment, and (3) with standard linear, additive covariate adjustment:

```{r}
## With the Lin Estimator
lin_adjust <- 
  
  rep_data %>%
  
  estimate(
    form = y ~ x + z1 + z2 + x:cz1 + x:cz2,
    vars = "x"
    ) %>%
  
  evaluate_power(
    delta = seq(0, 0.5, len = 200)
    )

## With no covariate adjustment
no_adjust <-
  
  rep_data %>%
  
  estimate(
    form = y ~ x,
    vars = "x"
    ) %>%
  
  evaluate_power(
    delta = seq(0, 0.5, len = 200)
    )

## With linear, additive covariate adjustment
standard_adjust <-
  
  rep_data %>%
  
  estimate(
    form = y ~ x + z1 + z2,
    vars = "x"
    ) %>%
  
  evaluate_power(
    delta = seq(0, 0.5, len = 200)
    )
```

We can now compare results under these alternative empirical strategies. Figure 3 shows the power curves for each approach. The Lin estimator provides substantial improvements in power over both no covariate adjustment and linear additive adjustment.^[This is, of course, true by design here. In real projects, the differences between these covariate adjustment strategies could often be smaller or negligible. But it is useful to remember that @lin_agnostic_2013 adjustment may be especially valuable with especially imbalanced designs and/or designs where covariates are likely correlated with treatment effect heterogeneity.] And it only took a few lines of code to get this result!

```{r}
## Combine the results, and apply similar plotting code to above
bind_rows(
  lin_adjust %>% mutate(Method = "Lin"),
  no_adjust %>% mutate(Method = "No Covariates"),
  standard_adjust %>% mutate(Method = "Additive")
  ) %>%
  
  ggplot() +
  
  geom_line(
    aes(delta, power, linetype = Method)
    ) +
  
  scale_linetype_manual(values = c("solid", "longdash", "dotted")) +
  
  geom_hline(
    yintercept = 0.8,
    color = "grey25",
    alpha = 0.8
    ) +
  
  scale_y_continuous(
    n.breaks = 6
    ) +
  
  labs(
    x = expression(delta),
    y = "Power",
    title = "Power with Lin Adjustment",
    linetype = "Method:"
    ) +
  
  ggridges::theme_ridges(
    font_size = 10,
    center_axis_labels = TRUE
    ) +
  
  theme(
    legend.position = "bottom"
    ) 
```

### Incorporating DeclareDesign into OES Power Tools

We can also use `DeclareDesign` within this **Replicate**, **Estimate**, **Evaluate** framework. This involves using `DeclareDesign` to draw estimates, and then feeding the results into the OES `evaluate_power()` function. We compare the `DeclareDesign` approach to the OES `Replicate` and `Estimate` steps below.

First, we simulate a simple design with the OES tools introduced above:

```{r}
eval <- 
  
  replicate_design(
    R = 1000,
    N = 100,
    Y = rnorm(N),
    Z = rbinom(N, 1, 0.5)
    ) %>%
  
  estimate(
    form = Y ~ Z, vars = "Z"
    ) %>%
  
  evaluate_power(
    delta = seq(0, 0.6, len = 10)
    )
```

Then, we do the same with `DeclareDesign`, declaring a population, potential outcomes, assignments, a target quantity of interest, and an estimator:

```{r}
design <-
  
  declare_population(
    N = 100,
    U = rnorm(N),
    potential_outcomes(Y ~ U)
    ) +
  
  declare_assignment(Z = simple_ra(N, prob = 0.5)) +
  
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  
  declare_estimator(
    Y ~ Z,
    inquiry = "ATE",
    .method = lm_robust
    )
```

We then use draws from this design within the OES tools:

```{r}
dd_eval <-
  
  replicate(
    n = 1000,
    expr = draw_estimates(design) %>% list
    ) %>%
  
  bind_rows() %>%
  
  evaluate_power(
    delta = seq(0, 0.6, len = 10)
    )
```

We show the similarity between the two approaches to generating the simulated
data in the figure below:

```{r}
bind_rows(
  eval %>% mutate(method = "OES Power Tools"),
  dd_eval %>% mutate(method = "DeclareDesign")
  ) %>%
  
  ggplot() +
  
  geom_line(
    aes(delta, power, linetype = method)
    ) +
  
  scale_linetype_manual(values = c("solid", "longdash")) +
  
  labs(
    x = expression(delta),
    y = "Power",
    linetype = NULL
    ) +
  
  scale_y_continuous(
    n.breaks = 6
    ) +
  
  geom_hline(
    yintercept = 0.8,
    col = "grey25",
    size = 1,
    alpha = 0.8
    ) +
  
  ggridges::theme_ridges(
    center_axis_labels = TRUE,
    font_size = 10
    ) +
  
  theme(
    legend.position = "bottom"
    )
```



