---
title: OES Standard Operating Procedures for The Design and Statistical Analysis of Experiments.
author: Jake Bowers, Ryan T. Moore, Lula Chen, Paul Testa, Nate Higgins, Oliver McClellan, Miles Williams, Tyler Simko, Bill Schultz
date: '`r format(Sys.Date(), "%B %d, %Y")`'
site: bookdown::bookdown_site
knit: bookdown::render_book
documentclass: book
bibliography: ["sop.bib", "packages.bib"]
link-citations: yes
colorlinks: yes
lot: yes
lof: yes
github-repo: gsa-oes/sop
description: "These are the current standard operating procedures for statistical analysis of the Office of Evaluation Sciences in the GSA"
fontsize: 12pt
geometry: margin=1in
graphics: yes
---


```{r options, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
require(knitr)
opts_chunk$set(strip.white=TRUE,
               width.cutoff=132,
               size='\\scriptsize',
               out.width='.9\\textwidth',
               message=FALSE,
               warning=FALSE,
               echo=TRUE,
               comment=NA,
               tidy='styler',
               prompt=FALSE,
               results='markup')
```

```{r loadlibs}
## Libraries are now managed by the renv system
library(bfe)
library(blockTools)
library(coin)
library(DeclareDesign)
library(devtools)
library(estimatr)
library(fabricatr)
library(foreach)
library(future)
library(future.apply)
library(here)
library(ICC)
library(kableExtra)
library(lmtest)
library(multcomp)
library(nbpMatching)
library(quickblock)
library(randomizr)
library(ri2)
library(sandwich)
library(tidyverse)
library(RItools)

options(
  htmltools.dir.version = FALSE, formatR.indent = 2,
  width = 100, digits = 4, warnPartialMatchAttr = FALSE, warnPartialMatchDollar = FALSE
)

local({
  r = getOption('repos')
  if (!length(r) || identical(unname(r['CRAN']), '@CRAN@'))
    r['CRAN'] = 'https://cran.rstudio.com'
  options(repos = r)
})
```


```{r htmlTemp3, echo=FALSE, eval=TRUE}
## This next from https://stackoverflow.com/questions/45360998/code-folding-in-bookdown
codejs <- readr::read_lines("js/codefolding.js")
collapsejs <- readr::read_lines("js/collapse.js")
transitionjs <- readr::read_lines("js/transition.js")

## Default to showing code
## window.initializeCodeFolding("show" === "show");
## Default to hiding code
## window.initializeCodeFolding("show" === "show");

htmlhead <-
  paste('
<script>',
paste(transitionjs, collapse = "\n"),
'</script>
<script>',
paste(collapsejs, collapse = "\n"),
'</script>
<script>',
paste(codejs, collapse = "\n"),
'</script>
<style type="text/css">
.code-folding-btn { margin-bottom: 4px; }
.row { display: flex; }
.collapse { display: none; }
.in { display:block }
</style>
<script>
$(document).ready(function () {
  window.initializeCodeFolding("show" === "hide");
});
</script>
', sep = "\n")

readr::write_lines(htmlhead, path = "header.html")
```

\newcommand{\var}{\mathrm{Var}}
\newcommand{\sd}{\mathrm{sd}}
\newcommand{\cov}{\mathrm{Cov}}
\newcommand{\cor}{\mathrm{Cor}}
\newcommand{\pr}{\text{Pr}}
\newcommand{\rank}{\text{rank}}
\newcommand{\Dt}{\Delta t}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bZ}{\mathbf{Z}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bE}{\mathbf{E}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bpsi}{\boldsymbol{\psi}}
\newcommand{\bmu}{\boldsymbol{\mu}}%m
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\btau}{\boldsymbol{\tau}}
\newcommand{\balpha}{\boldsymbol{\alpha}}
\newcommand{\bgamma}{\boldsymbol{\gamma}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}



# Overview {-}

This document explains **how** our team, the [Office of Evaluation Sciences in
the General Services Administration](https://oes.gsa.gov/) (the OES), tends to
do statistical analysis and it also explains **why** we do what we do. ^[We call
this document a standard operating procedure (SOP) because we are inspired by
the [Green, Lin and Coppock SOP](https://github.com/acoppock/Green-Lab-SOP).]
The research integrity processes OES follows is already documented on our [Evaluation Resources Web Page](https://oes.gsa.gov/methods/). For example, on that page we provide templates for our research design and analysis pre-registration process. Here, we get into the nitty gritty of our statistical work.

## Purposes of this document {-}

First, this document educates new team members about the decisions past team
members have made regarding the design and analysis of the studies fielded so
far. It also serves as a place to record decisions for [our
own future selves](http://dx.doi.org/10.4067/S0718-090X2016000300011), and helps us harness the power that arises from our disciplinary diversity. That
is, we have decided, as a team, how to do certain statistical analyses, and
these decisions may differ from those that are common in any given academic
discipline. This document, thus, helps explain why we have landed on those
decisions (for now), and how to implement those practices.

Second, this document records decisions that we have made in the absence of
pre-analysis plans, or in the context of circumstances unforeseen by our
pre-analysis planning. It should help guide our future experimental design
and analysis.

Third, this document should help us write better analysis plans and speed our
practice of re-analysis. (Our team insists on a blind re-analysis of every
study as a quality control for our results before they are reported to our
agency partners.)

Fourth, this document should help other teams working to learn about the causal impacts of policy interventions. We hope this contributes to the federal government's own work pursuing evidence-based public policy, but also helps other teams in other places doing work that is similar to our own.


## Nature and limitations of this document {-}

### We (mostly) focus on randomized field experiments. {-}

This document focuses on design and analysis of randomized field experiments.
Although we include some discussion about non-randomized studies, often known
as observational studies, until now, our team has focused primarily on
randomized field experiments. We plan to include more discussion of observational studies over time.

### We present examples using R {-}

As public servants and social and behavioral scientists, we use the [R](http://r-project.org) statistical analysis language for this document because it is (a) one of the two industry standards in the field of data science (along with Python), (b) free, open source, and multiplatform, and (c) the standard for advanced methodological work in the statistical sciences as applied to the social and behavioral sciences (the latest new statistical techniques for social and behavioral scientists tend to be developed in R). 

Many members of our team use Stata or SAS or SPSS or Python. We welcome
additions to this document using those languages as well.

### Structure of the document {-}

Each section of this document will include, if applicable:

 1. A description of our approach.
 2. A description of how we implement our approach, including functions in R,
    key arguments that must be entered into the function, and key
    values that are outputs from the function.
 3. A general example using simulated data (perhaps including some evaluation of the tool as compared to other possible choices).
 4. A discussion of a specific example from OES (if applicable) in which we implemented the given procedure.


Throughout the document, we include links to the [Glossary](#glossary) and
[Appendix](#appendix), to clarify terms or explain tools and procedures in
more depth.

## Help us improve our work! {-}

Since we hope to improve our analytic workflow with every project, this document should be seen as provisional --- as a record and a guide for our continuous learning and improvement.  We invite comments in the [Issues](https://github.com/gsa-oes/sop/issues) and [pull requests](https://help.github.com/articles/creating-a-pull-request/) for direct contributions.

<!-- This next copied from https://github.com/hadley/adv-r/blob/master/Introduction.Rmd -->

## About this document {-}

This book was written in [bookdown](http://bookdown.org/). 
The complete source is available from [GitHub](https://github.com/gsa-oes/sop). 
This version of the book was built with `r R.version.string` and the following packages.

```{r, echo = FALSE, results="asis"}
deps <- desc::desc_get_deps()$package[-1]
pkgs <- sessioninfo::package_info(deps, dependencies = FALSE)
df <- tibble(
  package = pkgs$package,
  version = pkgs$ondiskversion,
  source = gsub("@", "\\\\@", pkgs$source)
)
knitr::kable(df, format = "markdown")
```
```{r, include = FALSE}
ruler <- function(width = getOption("width")) {
  x <- seq_len(width)
  y <- case_when(
    x %% 10 == 0 ~ as.character((x %/% 10) %% 10),
    x %% 5 == 0  ~ "+",
    TRUE         ~ "-"
  )
  cat(y, "\n", sep = "")
  cat(x %% 10, "\n", sep = "")
}
ruler()
```

```{r makepackagesbib, include=FALSE, warnings=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(unique(c( df$package, 'bookdown', 'knitr', 'rmarkdown')), 'packages.bib')
```

