---
title: "Language button template"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Notice the numbers in the tab names: e.g., R0, rather than R. The tabs need unique names, as these names corresponding to HTML IDs (they don't function right if IDs are duplicated). Numbering each round of tabs by chapter seems easiest for organization, but anything unique to a code block is fine.

The chunk option `highlight = F` prevents a `code` button from appearing within the tabs; the CSS/Javascript used to implement code folding takes advantage of certain HTML tags that the highlight feature apparently uses as well. Alternatively, we could replace all show/hide buttons with tabs like this, including cases where there is only R code. This would let us drop the show/hide buttons entirely, and also allow syntax highlighting back inside the tabs.

::: {.tab} 
<button class="tablinks" onclick="unrolltab(event, 'ch0R0')">R code</button>
<button class="tablinks" onclick="unrolltab(event, 'ch0Stata0')">Stata code</button>
<button class="tablinks" onclick="unrolltab(event, 'ch0Hide0')">Hide</button>
::: {#ch0R0 .tabcontent} 
<br />
```{r, highlight = F, eval = F}
# R code here
```
:::
::: {#ch0Stata0 .tabcontent} 
<br />
```{stata, highlight = F, eval = F}
* Stata code here
```
::: 
::: {#ch0Hide0 .tabcontent}
::: 
:::

